Assessment of zonally symmetric and asymmetric components of the Southern Annular Mode using a novel approach

### 2021-07-08 analysis/paper/asymsam.Rmd
Mayúsculas

Positive trends in the SAM have been documented by various researchers using different indices, mostly on austral Summer and Autumn [e.g. @fogt2020 and references therein].
It is thought that these trends are driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables [@marshall2004; @gillett2005; @arblaster2006; @gillett2013].
However, it's not clear yet how or if the asymmetric SAM component responds to these forcings, or how its variability alters the observed trends.

@@

The linear trends for each of the indices (SAM, S\nobreakdash-SAM and A\nobreakdash-SAM) were evaluated for the complete period 1979 -- 2018 at each level for the whole year and separated by trimesters (Figure \@ref(fig:trends)).
The SAM index presents a statistically positive significant trend (Figure \@ref(fig:trends)`r get_tag(term == "full" & season == "Annual")`) that extends throughout the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
The seasonal trends (rest of Figure \@ref(fig:trends) column a) indicate that positive trends are present in Autumn and particularly in Summer, where the 100 hPa maximum is much more defined.
In Winter and Spring, we detect no statistically significant trend.
This is consistent with the results of previous studies, which find large positive trends in Summer, smaller in Autumn and no trends in the other seasons [e.g. @fogt2020 and references therein] using indices of the SAM based on surface or near-surface circulation.

By separating the SAM signal in its asymmetric and symmetric parts, not only we can see that these trends are almost entirely due to the symmetric component (column b vs. column c in Figure \@ref(fig:trends)), but in some cases the trends become clearer.
In Summer, A\nobreakdash-SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the SAM index; as a result, trends computed using only the Symmetric component are stronger (compare the shading region in Figure \@ref(fig:trends)`r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`).
In Autumn, the S\nobreakdash-SAM index reveals a statistically significant positive trend in the stratosphere that is not significant using the SAM index.

@@

Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at 700 hPa with surface land air and sea temperature for each trimester.
In Summer, positive values of the SAM index (Figure \@ref(fig:regr-air-season)`r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies in the mid latitudes.
The ring is not zonally symmetric, as there are four distinctive local maximums around 30\degree W, 120\degree W, 150\degree E and 90\degree E respectively.
In the tropics, there are negative anomalies in the equatorial Pacific, consistent with the negative correlation between SAM and ENSO.
Figure \@ref(fig:regr-air-season)`r combine_words(get_tag(season == "DJF" & term != "full", n = 2))` show temperature anomalies associated with positive values of A\nobreakdash-SAM and S\nobreakdash-SAM, respectively.
Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the A\nobreakdash-SAM regression map, while temperature patterns linked to positive S\nobreakdash-SAM show a more zonally consistent ring and less relation to the tropics.
Over Antarctica positive values of the SAM index are associated with negative temperature anomalies, particularly over the Eastern coast.
These anomalies are associated only with the S\nobreakdash-SAM.
Furthermore, the typical longitudinal variations of temperature anomalies along the Antarctic Peninsula are not evident in the regressions with the SAM, in agreement with previous works [e.g. @marshall2016].
On the other hand, temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to A\nobreakdash-SAM and are not present in the regression pattern with the SAM.

In Autumn, Winter and Spring (rows `r combine_words(get_row(season %in% c("MAM", "JJA", "SON"), n = 3))` in Figure \@ref(fig:regr-air-season)) the positive ring is only present through its local maximums in the regression with the SAM, which reflects the more asymmetric nature of the SAM compared to Summer.
There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
Similar features were observed in station measurements by @jones2019, although using data from 1957 to 2016.
In Spring, the tropical signal of A\nobreakdash-SAM is similar to the one in Summer, again revealing the significance of the ENSO-A\nobreakdash-SAM link. 
Moreover, from Autumn to Spring, positive (negative) temperature anomalies are discernible at the northern (southern) portion of the Antarctic Peninsula in the regressions with the SAM, which are also evident in the regressions with A-SAM. In agreement, @marshall2016 found the same signal in association with SAM while that pattern is only evident in Autumn for the southern baroclinic annular mode, which is associated with variations in extratropical storm amplitude.

@@

The pattern of negative anomalies in the pole surrounded by positive anomalies roughly seen in all seasons -- although with varying intensity and small-scale details -- translates to enhanced meridional temperature gradient maximised in the zero line, which is consistent with the intensification and poleward migration of the westerlies commonly linked to the SAM through thermal wind balance.
It's then not surprising to see it more clearly in association with S\nobreakdash-SAM (at least in Summer and Spring).
Temperatures over Eastern Antarctica are most impacted by the S\nobreakdash-SAM, while on Western Antarctica they are more sensitive to the A\nobreakdash-SAM.
Considering the positive trend in the S\nobreakdash-SAM index, this might be consistent with temperature trends, which show warming in the West and no trend in the East [@nicolas2014].

Figure \@ref(fig:regr-air-season) column b can be partially compared with Figure 11 from @fogt2012.
Although they used station data from 1958 to 2001, main features are reproduced here, such as the strong signal in New Zealand and Australia in Summer and Spring.

@@

The seasonal-based regressions show statistically significant anomalies only in Spring, when positive SAM is associated with positive precipitation anomalies in Eastern Australia (Figure \@ref(fig:pp-regr-oceania)a.5).
In this trimester, S\nobreakdash-SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the East Coast (Figure \@ref(fig:pp-regr-oceania)c.5) while the positive precipitation anomalies related to positive A\nobreakdash-SAM affect all Eastern Australia (Figure \@ref(fig:pp-regr-oceania)b.5).

In Summer, positive SAM index is associated with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (Figure \@ref(fig:pp-regr-oceania)a.2).
The Eastern part being dominated by the relationship with S\nobreakdash-SAM and the Western, by A\nobreakdash-SAM.
In Autumn, the regression with SAM shows positive anomalies in the North, similar to Summer, which are associated with the A\nobreakdash-SAM.
In Winter regression coefficients are much weaker than during the rest of the year. 
None of these regression coefficients are statistically significant at the 95% level.
The Spring signal is broadly consistent with @hendon2007, but whereas they also detected a strong signal in Summer, Figure \@ref(fig:pp-regr-oceania)a.2 shows no statistically significant association (although the coefficients have the consistent sign).

@@

Except Winter, seasonal-based regressions mirror this same pattern.
Even if not statistically significant, they all show negative values in SESA and southern Chile along with positive values in southern Brazil in relation with the SAM.
The separation of these features between A\nobreakdash-SAM and S\nobreakdash-SAM regression maps is also rather consistent.

@@

We observe positive trends of the SAM in Summer and Autumn, as was documented by previous studies [e.g. @fogt2020 and references therein] for low levels.
We show that these trends maximise at 100 hPa, and are explained by the zonally symmetric component.
We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not evident in the SAM index.
We find some evidence of the SAM becoming more zonally asymmetric during DJF, as there is a slight positive trend in the variance explained by the as the A\nobreakdash-SAM.
This is in contrast to @fogt2012 who made the evaluation for the period 1958 -- 2001.
This discrepancy might be due either due to differences in methodology or analysed period.

>>>>

Positive trends in the SAM have been documented by various researchers using different indices, mostly in austral summer and autumn [e.g. @fogt2020 and references therein].
It is thought that these trends are driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables [@marshall2004; @gillett2005; @arblaster2006; @gillett2013].
However, it's not clear yet how or if the asymmetric SAM component responds to these forcings, or how its variability alters the observed trends.

@@

The linear trends for each of the indices (SAM, S\nobreakdash-SAM and A\nobreakdash-SAM) were evaluated for the complete period 1979 -- 2018 at each level for the whole year and separated by trimesters (Figure \@ref(fig:trends)).
The SAM index presents a statistically positive significant trend (Figure \@ref(fig:trends)`r get_tag(term == "full" & season == "Annual")`) that extends throughout the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
The seasonal trends (rest of Figure \@ref(fig:trends) column a) indicate that positive trends are present in autumn and particularly in summer, where the 100 hPa maximum is much more defined.
In winter and spring, we detect no statistically significant trend.
This is consistent with the results of previous studies, which find large positive trends in summer, smaller in autumn and no trends in the other seasons [e.g. @fogt2020 and references therein] using indices of the SAM based on surface or near-surface circulation.

By separating the SAM signal in its asymmetric and symmetric parts, not only we can see that these trends are almost entirely due to the symmetric component (column b vs. column c in Figure \@ref(fig:trends)), but in some cases the trends become clearer.
In summer, A\nobreakdash-SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the SAM index; as a result, trends computed using only the Symmetric component are stronger (compare the shading region in Figure \@ref(fig:trends)`r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`).
In autumn, the S\nobreakdash-SAM index reveals a statistically significant positive trend in the stratosphere that is not significant using the SAM index.

@@

Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at 700 hPa with surface land air and sea temperature for each trimester.
In summer, positive values of the SAM index (Figure \@ref(fig:regr-air-season)`r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies in the mid latitudes.
The ring is not zonally symmetric, as there are four distinctive local maximums around 30\degree W, 120\degree W, 150\degree E and 90\degree E respectively.
In the tropics, there are negative anomalies in the equatorial Pacific, consistent with the negative correlation between SAM and ENSO.
Figure \@ref(fig:regr-air-season)`r combine_words(get_tag(season == "DJF" & term != "full", n = 2))` show temperature anomalies associated with positive values of A\nobreakdash-SAM and S\nobreakdash-SAM, respectively.
Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the A\nobreakdash-SAM regression map, while temperature patterns linked to positive S\nobreakdash-SAM show a more zonally consistent ring and less relation to the tropics.
Over Antarctica positive values of the SAM index are associated with negative temperature anomalies, particularly over the Eastern coast.
These anomalies are associated only with the S\nobreakdash-SAM.
Furthermore, the typical longitudinal variations of temperature anomalies along the Antarctic Peninsula are not evident in the regressions with the SAM, in agreement with previous works [e.g. @marshall2016].
On the other hand, temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to A\nobreakdash-SAM and are not present in the regression pattern with the SAM.

In autumn, winter and spring (rows `r combine_words(get_row(season %in% c("MAM", "JJA", "SON"), n = 3))` in Figure \@ref(fig:regr-air-season)) the positive ring is only present through its local maximums in the regression with the SAM, which reflects the more asymmetric nature of the SAM compared to summer.
There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
Similar features were observed in station measurements by @jones2019, although using data from 1957 to 2016.
In spring, the tropical signal of A\nobreakdash-SAM is similar to the one in summer, again revealing the significance of the ENSO-A\nobreakdash-SAM link. 
Moreover, from autumn to spring, positive (negative) temperature anomalies are discernible at the northern (southern) portion of the Antarctic Peninsula in the regressions with the SAM, which are also evident in the regressions with A-SAM. In agreement, @marshall2016 found the same signal in association with SAM while that pattern is only evident in autumn for the southern baroclinic annular mode, which is associated with variations in extratropical storm amplitude.

@@

The pattern of negative anomalies in the pole surrounded by positive anomalies roughly seen in all seasons -- although with varying intensity and small-scale details -- translates to enhanced meridional temperature gradient maximised in the zero line, which is consistent with the intensification and poleward migration of the westerlies commonly linked to the SAM through thermal wind balance.
It's then not surprising to see it more clearly in association with S\nobreakdash-SAM (at least in summer and spring).
Temperatures over Eastern Antarctica are most impacted by the S\nobreakdash-SAM, while on Western Antarctica they are more sensitive to the A\nobreakdash-SAM.
Considering the positive trend in the S\nobreakdash-SAM index, this might be consistent with temperature trends, which show warming in the West and no trend in the East [@nicolas2014].

Figure \@ref(fig:regr-air-season) column b can be partially compared with Figure 11 from @fogt2012.
Although they used station data from 1958 to 2001, main features are reproduced here, such as the strong signal in New Zealand and Australia in summer and spring.

@@

The seasonal-based regressions show statistically significant anomalies only in spring, when positive SAM is associated with positive precipitation anomalies in Eastern Australia (Figure \@ref(fig:pp-regr-oceania)a.5).
In this trimester, S\nobreakdash-SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the East Coast (Figure \@ref(fig:pp-regr-oceania)c.5) while the positive precipitation anomalies related to positive A\nobreakdash-SAM affect all Eastern Australia (Figure \@ref(fig:pp-regr-oceania)b.5).

In summer, positive SAM index is associated with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (Figure \@ref(fig:pp-regr-oceania)a.2).
The Eastern part being dominated by the relationship with S\nobreakdash-SAM and the Western, by A\nobreakdash-SAM.
In autumn, the regression with SAM shows positive anomalies in the North, similar to summer, which are associated with the A\nobreakdash-SAM.
In winter regression coefficients are much weaker than during the rest of the year. 
None of these regression coefficients are statistically significant at the 95% level.
The spring signal is broadly consistent with @hendon2007, but whereas they also detected a strong signal in summer, Figure \@ref(fig:pp-regr-oceania)a.2 shows no statistically significant association (although the coefficients have the consistent sign).

@@

Except winter, seasonal-based regressions mirror this same pattern.
Even if not statistically significant, they all show negative values in SESA and southern Chile along with positive values in southern Brazil in relation with the SAM.
The separation of these features between A\nobreakdash-SAM and S\nobreakdash-SAM regression maps is also rather consistent.

@@

We observe positive trends of the SAM in summer and autumn, as was documented by previous studies [e.g. @fogt2020 and references therein] for low levels.
We show that these trends maximise at 100 hPa, and are explained by the zonally symmetric component.
We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not evident in the SAM index.
We find some evidence of the SAM becoming more zonally asymmetric during DJF, as there is a slight positive trend in the variance explained by the as the A\nobreakdash-SAM.
This is in contrast to @fogt2012 who made the evaluation for the period 1958 -- 2001.
This discrepancy might be due either due to differences in methodology or analysed period.
